---
title: "Chapter 4 Homework Solutions"
author: "Ali Raisolsadat"
date: "2026-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Chapter 4 - Lecture 1


```{r, out.width="100%"}
target_density <- function(x) {
  # target density for the Metropolis algorithm
  return(exp(-0.5 * x^2))
}

metropolis_step <- function(p, x, proposal_std) {
  # propose a new state using Gaussian random walk
  x_hat <- x + rnorm(1, mean = 0, sd = proposal_std)

  # compute acceptance probability
  alpha <- min(1, p(x_hat) / p(x))

  # accept or reject the proposal
  u <- runif(1, 0, 1)

  if (u < alpha) {
    x <- x_hat  # accept
  }

  return(x)
}

# initialize a single chain
set.seed(0)
n_steps <- 200
proposal_std <- 2
x <- numeric(n_steps)
x[1] <- -10

# simulate a single chain
for (t in 2:n_steps) {
  x[t] <- metropolis_step(target_density, x[t-1], proposal_std)
}

# plot
# increase margins for large figure
par(mar = c(5,5,4,2))
plot(x,type = "l", lwd = 1, xlab = "Iteration", ylab = "x", main = "Burn-in and Thinning")
usr <- par("usr")
rect(xleft = 0, ybottom = usr[3], xright = 30, ytop = usr[4], col = rgb(0.5, 0.5, 0.5, 0.3), border = NA)
points(seq(31, n_steps, by = 5), x[seq(31, n_steps, by = 5)], col = "red", pch = 16)
legend("bottomright", legend = c("Chain", "Burn-in", "Thinned samples"), 
       col = c("black", rgb(0.5,0.5,0.5,0.3), "red"), 
       lty = c(1, NA, NA), pch = c(NA, 15, 16), pt.cex = 2, bty = "n"
)
```

```{r, out.width="100%"}
# simulation parameters
n_chains <- 1000       # number of independent chains
burn_in <- 30
thin <- 5
chains <- runif(n_chains, -10, 10)

# run chains and collect thinned samples
thinned_samples <- c()

for (t in 1:n_steps) {
  for (i in 1:n_chains) {
    chains[i] <- metropolis_step(target_density, chains[i], proposal_std)
  }
  
  # after burn-in, collect thinned samples
  if (t > burn_in && (t - burn_in) %% thin == 0) {
    thinned_samples <- c(thinned_samples, chains)
  }
}

# plot histogram of thinned samples
hist(thinned_samples, breaks = 50, probability = TRUE, col = "steelblue", border = "gray",
     main = paste("Histogram of Thinned Samples from", n_chains, "Independent Chains"),
     xlab = "x", ylab = "Density"
)
grid()
```

### Comment on the Histogram

After running many independent chains, applying burn-in and thinning, the combined thinned samples form a histogram that approximates the target distribution.

The histogram shows the probability density of the target, with higher frequencies where the distribution has higher probability (near the mean, lower spread in the tails).

As the number of chains and samples increases, the histogram converges to the true Normal(0,10) density.

This demonstrates the key property of MCMC: even though individual samples are dependent and start from different points, after burn-in and with sufficient sampling, the collected samples are effectively drawn from the stationary distribution.

### Chapter 4 Homework - Lecture 2
```{r, out.width="100%"}

# target distribution
target <- function(x) {
  return(exp(-x^2 / 8))
}

# asymmetric proposal mechanism for Metropolis Hastings
proposal_fn <- function(x, step_size = 1.0) {
  if (runif(1) < 0.7) {
    return(rnorm(1, mean = x + step_size, sd = 1))
  } else {
    return(rnorm(1, mean = x - step_size, sd = 1))
  }
}

# evaluate proposal density q(x_new | x_old)
proposal_density_fn <- function(x_new, x_old, step_size = 1.0) {
  return(
    0.7 * dnorm(x_new, mean = x_old + step_size, sd = 0.5) +
    0.3 * dnorm(x_new, mean = x_old - step_size, sd = 0.5)
  )
}

# Metropolis Hastings algorithm
metropolis_hastings <- function(target, T = 5000, x0 = 0.0, 
                                proposal_fn = NULL, proposal_density_fn = NULL, step_size = 0.8) {
  # initialize chain storage
  x <- numeric(T + 1)
  x[1] <- x0

  for (t in 2:(T + 1)) {
    # propose a new candidate
    x_prop <- proposal_fn(x[t - 1], step_size)

    # compute acceptance ratio
    numerator <- target(x_prop)
    denominator <- target(x[t - 1])

    # if asymmetric proposal include ratio
    if (!is.null(proposal_density_fn)) {
      numerator <- numerator *
        proposal_density_fn(x[t - 1], x_prop, step_size)

      denominator <- denominator *
        proposal_density_fn(x_prop, x[t - 1], step_size)
    }

    alpha <- min(1, numerator / denominator)

    # acceptance condition
    if (runif(1) < alpha) {
      x[t] <- x_prop
    } else {
      x[t] <- x[t - 1]
    }
  }

  return(x)
}


# MCMC settings
T <- 2000
burn_in <- 40
thinning <- 5


# run a single Metropolis–Hastings chain
x_single <- metropolis_hastings(target = target,T = T,x0 = 0,
                                proposal_fn = proposal_fn, 
                                proposal_density_fn = proposal_density_fn, 
                                step_size = 1.0
)


# burn-in and thinning
post_burnin <- x_single[(burn_in + 1):length(x_single)]
thin_idx <- seq(1, length(post_burnin), by = thinning)
thinned_samples <- post_burnin[thin_idx]
thinned_iters <- thin_idx + burn_in


# plot
plot(x_single,type = "l",lwd = 1, main = "Metropolis–Hastings Chain", xlab = "Iteration", ylab = "State value")

# burn-in shading
usr <- par("usr")
rect(xleft = 0, ybottom = usr[3], xright = burn_in, ytop = usr[4], col = rgb(0.5, 0.5, 0.5, 0.3), border = NA)
points(thinned_iters, thinned_samples, col = "red", pch = 16, cex = 0.5)
legend("topright", legend = c("Markov chain", "Burn-in", "Thinned samples"),
       col = c("black", rgb(0.5,0.5,0.5,0.3), "red"),
       lty = c(1, NA, NA), pch = c(NA, 15, 16), pt.cex = 1, bty = "n"
)
grid()
```

```{r, out.width="100%"}
# multiple independent MCMC chains
num_chains <- 500              # number of chains to simulate
start_range <- c(-4, 8)        # range for random initialization (overdispersed)
all_samples <- numeric(0)

for (i in 1:num_chains) {
  x0 <- runif(1, start_range[1], start_range[2])
  chain <- metropolis_hastings(target, T, x0, proposal_fn,proposal_density_fn, 1.0)
  post_burnin <- chain[(burn_in + 1):length(chain)]
  all_samples <- c(all_samples, post_burnin[seq(1, length(post_burnin), by = thinning)])
}

# plot histogram
hist(all_samples, breaks = 50, probability = TRUE, col = "skyblue", border = "black",
  main = "Histogram of Combined Thinned Samples from 500 Chains", 
  xlab = "x", ylab = "Density"
)

# overlay true density N(0,4)
x_vals <- seq(-8, 8, length.out = 500)
lines(x_vals, dnorm(x_vals, mean = 0, sd = 2), col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("MCMC Samples", "Target N(0,4)"),
       fill = c("skyblue", NA), border = c("black", NA), lty = c(NA, 2), col = c(NA, "red"), bty = "n"
)

grid()
```
